%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% chapter2.tex
%% NOVA thesis document file
%%
%% Chapter with the template manual
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE chapter2.tex}%

\chapter{Related Work}
\label{cha:related_work}

\glsresetall

\section{VR Locomotion Techniques}
\label{sec:vr-locomotion-techniques}

Boletsis, C., \& Cedergren, J. E. (2019). VR Locomotion in the New Era of Virtual Reality: An Empirical Comparison of Prevalent Techniques. Advances in Human-Computer Interaction, 2019. https://doi.org/10.1155/2019/7420781

Boletsis, C. (2017). The new era of virtual reality locomotion: A systematic literature review of techniques and a proposed typology. In Multimodal Technologies and Interaction (Vol. 1, Issue 4). MDPI AG. https://doi.org/10.3390/mti1040024

\subsection{Joysticks and Teleportation}
\label{sec:joysticks-and-teleportation}

Langbehn, E., Lubos, P., \& Steinicke, F. (2018, April 4). Evaluation of locomotion techniques for room-scale VR: Joystick, teleportation, and redirected walking. ACM International Conference Proceeding Series. https://doi.org/10.1145/3234253.3234291

\subsection{Natural Walking}
\label{sec:natural-walking}

Cao, A., Wang, L., Liu, Y., \& Popescu, V. (2020). Feature Guided Path Redirection for VR Navigation. 137–145. https://doi.org/10.1109/vr46266.2020.00032

Exploring Large Virtual Environments with an HMD when Physical Space is Limited (2007)

Flexible Spaces: Dynamic Layout Generation for Infinite Walking in Virtual Environments . (2013). IEEE.

Suma, E. A., Lipps, Z., Finkelstein, S., Krum, D. M., \& Bolas, M. (2012). Impossible Spaces: Maximizing Natural Walking in Virtual Environments with Self-Overlapping Architecture.

Nakatsu, Ryohei., Klinker, Gudrun., \& Lok, Benjamin. (2010)

\subsubsection{Through Redirection}
\label{sec:redirection}

15 Years of Research on Redirected Walking in Immersive Virtual Environments. (2018). www.computer.org/cga

\subsubsection{Through Adaptable Spaces}
\label{sec:adaptable-spaces} 

Simeone, A. L., Christian Nilsson, N., Zenner, A., Speicher, M., \& Daiber, F. (2020). The Space Bender: Supporting Natural Walking via Overt Manipulation of the Virtual Environment. 598–606. https://doi.org/10.1109/vr46266.2020.00082

Suma, E. A., Lipps, Z., Finkelstein, S., Krum, D. M., \& Bolas, M. (2012). Impossible Spaces: Maximizing Natural Walking in Virtual Environments with Self-Overlapping Architecture.

\section{Non-Euclidean and Impossible Spaces}
\label{sec:non-euclidean-and-impossible-spaces}

\subsection{Non-Euclidean Geometry}
\label{sec:non-euclidean-geometry}

Rita, Rebelo, O. (2020). VR LAB: USER INTERACTION IN VIRTUAL ENVIRONMENTS USING SPACE AND TIME MORPHING.

Warren, W. H. (2019). Non-Euclidean navigation. In Journal of Experimental Biology (Vol. 222). Company of Biologists Ltd. https://doi.org/10.1242/jeb.187971

Hart, V., Hawksley, A., Matsumoto, E. A., \& Segerman, H. (2017). Non-euclidean Virtual Reality I: Explorations of H 3. In Conference Proceedings (Vol. 33).

Muryy, A., \& Glennerster, A. (2021). Route selection in non-Euclidean virtual environments. PLoS ONE, 16(4 April). https://doi.org/10.1371/journal.pone.0247818

%%Kopczy'nski, E. K., Celi'nska, D. C., & Marekčtrnáct, M. M. (2017). HyperRogue: Playing with Hyperbolic Geometry.

\subsection{Impossible Spaces}
\label{sec:impossible-spaces}

Suma, E. A., Lipps, Z., Finkelstein, S., Krum, D. M., \& Bolas, M. (2012). Impossible Spaces: Maximizing Natural Walking in Virtual Environments with Self-Overlapping Architecture.

\section{Space Modification Techniques}
\label{sec:spatial-techniques}

\subsection{Procedural Content Generation}
\label{sec:pcg}

Dong, Z. C., Fu, X. M., Zhang, C., Kang, W. U., \& Ligang, L. I. U. (2017). Smooth assembled mappings for large-scale real walking. ACM Transactions on Graphics, 36(6). https://doi.org/10.1145/3130800.3130893

Sra, M., Garrido-Jurado, S., \& Maes, P. (2018). Oasis: Procedurally Generated Social Virtual Spaces from 3D Scanned Real Spaces. IEEE Transactions on Visualization and Computer Graphics, 24(12), 3174–3187. https://doi.org/10.1109/TVCG.2017.2762691

\subsection{Spatial Compression}
\label{sec:spatial-compression}

Simeone, A. L., Christian Nilsson, N., Zenner, A., Speicher, M., \& Daiber, F. (2020). The Space Bender: Supporting Natural Walking via Overt Manipulation of the Virtual Environment. 598–606. https://doi.org/10.1109/vr46266.2020.00082

Towards Efficient Spatial Compression in Self-Overlapping Virtual Environments (2017)